Variable Selection in Convex Function Estimation

### 2013-11-05 tex/scam/appendix.tex
massive journal edit

\subsection{Proof of Additive Faithfulness}
\label{sec:faithful_proof}

We start with a proof of Lemma~\ref{lem:general_int_reduction}. We restate the lemma here for convenience.

\begin{lemma}
Let $F$ be a product distribution on $C=[0,1]^s$ with a positive density function $p$. Let $f: C \rightarrow \R$ be an integrable function.

Let $\ds f^*_1,...,f^*_s, \mu^* \coloneqq \arg\min \{ \E | f(X) - \sum_{k=1}^s f_k(X_k) - \mu |^2 \,:\, \forall k, \E f_k(X_k) = 0\}$

Then $f^*_k(x_k) = \E[ f(X) \given x_k] - \E f(X)$ and $\mu^* = \E f(X)$ and this solution is unique.
\end{lemma}

Let $f^*_1,...,f^*_s, \mu^*$ be the minimizers as defined.

We first show that the optimal $\mu^* = \E f(X)$ for any $f_1, ..., f_k$ such that $\E f_k(X_k) = 0$. This follows from the stationarity condition, which states that $\mu^* = \E[ f(X) - \sum_k f_k(X_k)] = \E[ f(X) ]$. \\

We now turn our attention toward the $f^*_k$'s.

It must be then that $f^*_k$ minimizes $\{ \E | f(X) - \mu^* - \sum_{k' \neq k} f^*_{k'} (X_{k'}) - f_k (X_k)|^2 \,:\, f_k \trm{ convex }, \E f_k(X_k) = 0 \}$.

Fix $x_k$, we will show that the value $\E[ f(X) | x_k] - \mu^*$, for all $x_k$,  uniquely minimizes 
\[
\min_{ f_k(x_k) } \int_{\mathbf{x}_{-k}} p(\mathbf{x}) 
         \big| f(\mathbf{x}) - \sum_{k' \neq k} f^*_{k'} (x_{k'}) - f_k (x_k) -\mu^*\big|^2 
                 d \mathbf{x}_{-k}.
\]
It easily follows then that the function $x_k \mapsto \E[ f(X) \given x_k] - \mu^*$ is the unique $f^*_k$ that minimizes the expected square error.

Take the derivative with respect to the value $f_k(x_k)$ and set it equal to zero, we get that
\begin{align*}
\int_{\mathbf{x}_{-k}} p(\mathbf{x}) f_k(x_k) d \mathbf{x}_{-k} &= 
  \int_{\mathbf{x}_{-k}} p(\mathbf{x}) 
      ( f(\mathbf{x})-\sum_{k' \neq k} f^*_{k'}(x_{k'})-\mu^*) d \mathbf{x}_{-k} \\  
p(x_k) f_k(x_k) &= \int_{\mathbf{x}_{-k}} p(x_k)
     p(\mathbf{x}_{-k}) ( f(\mathbf{x}) - \sum_{k' \neq k} f^*_{k'} (x_{k'})-\mu^*) 
              d \mathbf{x}_{-k} \\
f_k(x_k) &= \int_{\mathbf{x}_{-k}} 
       p(\mathbf{x}_{-k}) (f(\mathbf{x}) -\mu^*) d \mathbf{x}_{-k} 
 \end{align*}
Where the second line follows because $p(\mathbf{x}) = \prod_k p(x_k)$ by product distribution assumption. The third line follows because $\int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) f^*_{k'}(x_{k'}) d\mathbf{x}_{-k} = 0$ for all $k' \neq k$ by the constraint that $\E[f^*_{k'}(X_{k'})] = 0$.

The square error objective is strongly convex. The second derivative with respect to $f_k(x_k)$ is $2 p(x_k)$, which is always positive under the assumption that $p$ is positive. Therefore, the solution $f^*_k(x_k) = \E[ f(X) \given x_k ] - \E f(X)$ is unique.

Now, we note that as a function of $x_k$, $\E[ f(X) | x_k] - \E f(X)$ has mean zero and we thus finish the proof.

Our reasoning actually proves something stronger. It proves the following lemma:

\begin{lemma}
\label{lem:additive_component}

$f^*_k(x_k) = \E[ f(X) | x_k] - \mu^*$ is the unique optimal solution to 
\[
min_{f_k} \{ \E\Big( f(X) - \E f(X) - \sum_{k'\neq k} f_{k'}(X_{k'}) - f_k (X_k)\Big)^2 
                    \,:\, \E f_k(X_k) = 0 \}
\]
for \emph{any} set of component functions $\{ f_{k'} \}_{k'\neq k}$ that satisfies $\E f_{k'}(X_{k'}) = 0$.
\end{lemma}

The next lemma adapts Lemma~\ref{lem:general_int_reduction} to convex functions.

\begin{lemma}
\label{lem:int_reduction}
Let $F$ be a product distribution on $C=[0,1]^s$ with positive density function $p$. Let $f: C \rightarrow \R$ be a convex function.

Let $\ds f^*_1,...,f^*_s, \mu^* \coloneqq \arg\min \{ \E | f(X) - \sum_{k=1}^s f_k(X_k) - \mu |^2 \,:\, \forall k, f_k \trm{ convex }, \E f_k(X_k) = 0\}$

Then $f^*_k(x_k) = \E[ f(X) \given x_k] - \E f(X)$ and $\mu^* = \E f(X)$ and this solution is unique.
\end{lemma}

\begin{proof}
The proof is identical to that of Lemma~\ref{lem:general_int_reduction}. We need only verify that $f^*_k(x_k) = \E[ f(X) \given x_k] - \E f(X)$ is convex.

This is true because for every $\mathbf{x}_{-k}$, $f(x_k, \mathbf{x}_{-k})$ is a convex function with respect to $x_k$ and therefore, $\int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) f(x_k, \mathbf{x}_{-k}) d \mathbf{x}_{-k}$ is still convex. 
\end{proof}

 The next proposition, combined with Lemma~\ref{lem:int_reduction}, proves Theorem~\ref{thm:convex_faithful}.

\begin{proposition}
\label{prop:faithful_condition}
Let $F$ be a product distribution supported on $C=[0,1]^s$ with positive density $p > 0$. Let $f: C \rightarrow \R$ be a convex function, twice differentiable.

Let $f^*_1,...,f^*_s \coloneqq \arg\min \{ \E | f(X) -\E f(X) - \sum_k f_k(X_k) |^2 \,:\, \forall k, f_k \trm{ convex }, \E f_k(X_k) = 0 \}$.

The following are equivalent:
\begin{packed_enum}
\item $f$ does not depends on coordinate $k$
\item For all $x_k$, $\E[ f(X) | x_k] = \E f(X) $.
\end{packed_enum}
\end{proposition}

The first condition trivially implies the second.

Fix $k$. Suppose that, for all $x_k$, $\E[ f(X) | x_k] = \E f(X)$.

Because we assume that $p(\mathbf{x}) = \prod_k p(x_k)$, we see that, for all $x_k$,
\begin{align*}
 \E[ f(X) | x_k ] &=  \int_{\mathbf{x}_{-k}} p(\mathbf{x_{-k}}) f(x_k,\mathbf{x}_{-k}) d \mathbf{x}_{-k} = \E f(X)
\end{align*}

For every $\mathbf{x}_{-k}$, we define the derivative
\[
g(\mathbf{x}_{-k}) \coloneqq  \lim_{x_k \rightarrow 0^+} \frac{f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k})}{x_k}
\]
$g(\mathbf{x}_{-k})$ is well-defined by the assumption that $f$ is everywhere differentiable.

We now describe two facts about $g$.

Fact 1. By exchanging limit with the integral, which is valid by bounded convergence theorem, we reason that 
\[
\int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) g(\mathbf{x}_{-k}) d \mathbf{x}_{-k} = 0
\]

Fact 2. Because $f$ is convex, $g(\mathbf{x}_{-k})$ is a component of the subgradient $\partial_{\mathbf{x}} f(0, \mathbf{x}_{-k})$. (the subgradient coincides with the gradient by assumption that $f$ is twice differentiable)

Therefore, using the first order characterization of a convex function, we have
\begin{align*}
f(\mathbf{x}') &\geq f( \mathbf{x}) + \partial_{\mathbf{x}} f(\mathbf{x})^\tran (\mathbf{x}' - \mathbf{x}) \quad \trm{ for all } \mathbf{x}', \mathbf{x} \\
f(x_k, \mathbf{x}_{-k}) &\geq f(0, \mathbf{x}_{-k}) + g(\mathbf{x}_{-k}) x_k \quad \trm{ for all } x_k, \mathbf{x}_{-k}
\end{align*}

For all $x_k, \mathbf{x}_{-k}$, 
\[
f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k \geq 0
\]
and
\[
\int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) (f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k) d\mathbf{x}_{-k} = 0
\]
Since $f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k$ is a continuous function of $\mathbf{x}_{-k}$ and the density $p(\mathbf{x}_{-k})$ is, by assumption, non-zero on the support, we conclude that for all $x_k, f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k =0 $ necessarily and thus for all $\mathbf{x}_{-k}$, $f(x_k, \mathbf{x}_{-k}) = f(0, \mathbf{x}_{-k}) + g(\mathbf{x}_{-k}) x_k$.

The Hessian of $f$ then (guaranteed to exist by assumption) has a zero on the $k$-th main diagonal entry.

By proposition 7.1.10 from Horn and Johnson~\cite{HJ90},  such a matrix is positive semidefinite if and only if the $k$-th row and column are also zero. (to prove this, one can first work with a 2 by 2 matrix and then use the fact that every principal submatrix of a PSD matrix must be PSD)

Since $k$-th row and column correspond precisely to the gradient of $g(\mathbf{x}_{-k})$, we conclude that $g$ must be a constant function. It follows therefore that $g = 0$ because it integrates to 0.

So we have that for all $x_k, \mathbf{x}_{-k}$, $f(x_k, \mathbf{x}_{-k}) = f(0, \mathbf{x}_{-k})$, which concludes our proof.

@@

%%% Local Variables: ***
%%% mode:latex ***
%%% TeX-master: "icml14/scam_icml.tex" ***
%%% End: ***

>>>>

% \subsection{Proof of Additive Faithfulness}
% \label{sec:faithful_proof}

% We start with a proof of Lemma~\ref{lem:general_int_reduction}. We restate the lemma here for convenience.

% \begin{lemma}
% Let $F$ be a product distribution on $C=[0,1]^s$ with a positive density function $p$. Let $f: C \rightarrow \R$ be an integrable function.

% Let $\ds f^*_1,...,f^*_s, \mu^* \coloneqq \arg\min \{ \E | f(X) - \sum_{k=1}^s f_k(X_k) - \mu |^2 \,:\, \forall k, \E f_k(X_k) = 0\}$

% Then $f^*_k(x_k) = \E[ f(X) \given x_k] - \E f(X)$ and $\mu^* = \E f(X)$ and this solution is unique.
% \end{lemma}

%  \begin{proof}

% Let $f^*_1,...,f^*_s, \mu^*$ be the minimizers as defined.

% We first show that the optimal $\mu^* = \E f(X)$ for any $f_1, ..., f_k$ such that $\E f_k(X_k) = 0$. This follows from the stationarity condition, which states that $\mu^* = \E[ f(X) - \sum_k f_k(X_k)] = \E[ f(X) ]$. \\

% We now turn our attention toward the $f^*_k$'s.

% It must be then that $f^*_k$ minimizes $\{ \E | f(X) - \mu^* - \sum_{k' \neq k} f^*_{k'} (X_{k'}) - f_k (X_k)|^2 \,:\, f_k \trm{ convex }, \E f_k(X_k) = 0 \}$.

% Fix $x_k$, we will show that the value $\E[ f(X) | x_k] - \mu^*$, for all $x_k$,  uniquely minimizes 
% \[
% \min_{ f_k(x_k) } \int_{\mathbf{x}_{-k}} p(\mathbf{x}) 
%          \big| f(\mathbf{x}) - \sum_{k' \neq k} f^*_{k'} (x_{k'}) - f_k (x_k) -\mu^*\big|^2 
%                  d \mathbf{x}_{-k}.
% \]
% It easily follows then that the function $x_k \mapsto \E[ f(X) \given x_k] - \mu^*$ is the unique $f^*_k$ that minimizes the expected square error.

% Take the derivative with respect to the value $f_k(x_k)$ and set it equal to zero, we get that
% \begin{align*}
% \int_{\mathbf{x}_{-k}} p(\mathbf{x}) f_k(x_k) d \mathbf{x}_{-k} &= 
%   \int_{\mathbf{x}_{-k}} p(\mathbf{x}) 
%       ( f(\mathbf{x})-\sum_{k' \neq k} f^*_{k'}(x_{k'})-\mu^*) d \mathbf{x}_{-k} \\  
% p(x_k) f_k(x_k) &= \int_{\mathbf{x}_{-k}} p(x_k)
%      p(\mathbf{x}_{-k}) ( f(\mathbf{x}) - \sum_{k' \neq k} f^*_{k'} (x_{k'})-\mu^*) 
%               d \mathbf{x}_{-k} \\
% f_k(x_k) &= \int_{\mathbf{x}_{-k}} 
%        p(\mathbf{x}_{-k}) (f(\mathbf{x}) -\mu^*) d \mathbf{x}_{-k} 
%  \end{align*}
% Where the second line follows because $p(\mathbf{x}) = \prod_k p(x_k)$ by product distribution assumption. The third line follows because $\int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) f^*_{k'}(x_{k'}) d\mathbf{x}_{-k} = 0$ for all $k' \neq k$ by the constraint that $\E[f^*_{k'}(X_{k'})] = 0$.

% The square error objective is strongly convex. The second derivative with respect to $f_k(x_k)$ is $2 p(x_k)$, which is always positive under the assumption that $p$ is positive. Therefore, the solution $f^*_k(x_k) = \E[ f(X) \given x_k ] - \E f(X)$ is unique.

% Now, we note that as a function of $x_k$, $\E[ f(X) | x_k] - \E f(X)$ has mean zero and we thus finish the proof.

%  \end{proof}

% Our reasoning actually proves something stronger. It proves the following lemma:

% \begin{lemma}
% \label{lem:additive_component}

% $f^*_k(x_k) = \E[ f(X) | x_k] - \mu^*$ is the unique optimal solution to 
% \[
% min_{f_k} \{ \E\Big( f(X) - \E f(X) - \sum_{k'\neq k} f_{k'}(X_{k'}) - f_k (X_k)\Big)^2 
%                     \,:\, \E f_k(X_k) = 0 \}
% \]
% for \emph{any} set of component functions $\{ f_{k'} \}_{k'\neq k}$ that satisfies $\E f_{k'}(X_{k'}) = 0$.
% \end{lemma}

% The next lemma adapts Lemma~\ref{lem:general_int_reduction} to convex functions.

% \begin{lemma}
% \label{lem:int_reduction}
% Let $F$ be a product distribution on $C=[0,1]^s$ with positive density function $p$. Let $f: C \rightarrow \R$ be a convex function.

% Let $\ds f^*_1,...,f^*_s, \mu^* \coloneqq \arg\min \{ \E | f(X) - \sum_{k=1}^s f_k(X_k) - \mu |^2 \,:\, \forall k, f_k \trm{ convex }, \E f_k(X_k) = 0\}$

% Then $f^*_k(x_k) = \E[ f(X) \given x_k] - \E f(X)$ and $\mu^* = \E f(X)$ and this solution is unique.
% \end{lemma}

% \begin{proof}
% The proof is identical to that of Lemma~\ref{lem:general_int_reduction}. We need only verify that $f^*_k(x_k) = \E[ f(X) \given x_k] - \E f(X)$ is convex.

% This is true because for every $\mathbf{x}_{-k}$, $f(x_k, \mathbf{x}_{-k})$ is a convex function with respect to $x_k$ and therefore, $\int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) f(x_k, \mathbf{x}_{-k}) d \mathbf{x}_{-k}$ is still convex. 
% \end{proof}

%  The next proposition, combined with Lemma~\ref{lem:int_reduction}, proves Theorem~\ref{thm:convex_faithful}.

% \begin{proposition}
% \label{prop:faithful_condition}
% Let $F$ be a product distribution supported on $C=[0,1]^s$ with positive density $p > 0$. Let $f: C \rightarrow \R$ be a convex function, twice differentiable.

% Let $f^*_1,...,f^*_s \coloneqq \arg\min \{ \E | f(X) -\E f(X) - \sum_k f_k(X_k) |^2 \,:\, \forall k, f_k \trm{ convex }, \E f_k(X_k) = 0 \}$.

% The following are equivalent:
% \begin{packed_enum}
% \item $f$ does not depends on coordinate $k$
% \item For all $x_k$, $\E[ f(X) | x_k] = \E f(X) $.
% \end{packed_enum}
% \end{proposition}

% \begin{proof}

% The first condition trivially implies the second.

% Fix $k$. Suppose that, for all $x_k$, $\E[ f(X) | x_k] = \E f(X)$.

% Because we assume that $p(\mathbf{x}) = \prod_k p(x_k)$, we see that, for all $x_k$,
% \begin{align*}
%  \E[ f(X) | x_k ] &=  \int_{\mathbf{x}_{-k}} p(\mathbf{x_{-k}}) f(x_k,\mathbf{x}_{-k}) d \mathbf{x}_{-k} = \E f(X)
% \end{align*}

% For every $\mathbf{x}_{-k}$, we define the derivative
% \[
% g(\mathbf{x}_{-k}) \coloneqq  \lim_{x_k \rightarrow 0^+} \frac{f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k})}{x_k}
% \]
% $g(\mathbf{x}_{-k})$ is well-defined by the assumption that $f$ is everywhere differentiable.

% We now describe two facts about $g$.

% Fact 1. By exchanging limit with the integral, which is valid by bounded convergence theorem, we reason that 
% \[
% \int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) g(\mathbf{x}_{-k}) d \mathbf{x}_{-k} = 0
% \]

% Fact 2. Because $f$ is convex, $g(\mathbf{x}_{-k})$ is a component of the subgradient $\partial_{\mathbf{x}} f(0, \mathbf{x}_{-k})$. (the subgradient coincides with the gradient by assumption that $f$ is twice differentiable)

% Therefore, using the first order characterization of a convex function, we have
% \begin{align*}
% f(\mathbf{x}') &\geq f( \mathbf{x}) + \partial_{\mathbf{x}} f(\mathbf{x})^\tran (\mathbf{x}' - \mathbf{x}) \quad \trm{ for all } \mathbf{x}', \mathbf{x} \\
% f(x_k, \mathbf{x}_{-k}) &\geq f(0, \mathbf{x}_{-k}) + g(\mathbf{x}_{-k}) x_k \quad \trm{ for all } x_k, \mathbf{x}_{-k}
% \end{align*}

% For all $x_k, \mathbf{x}_{-k}$, 
% \[
% f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k \geq 0
% \]
% and
% \[
% \int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) (f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k) d\mathbf{x}_{-k} = 0
% \]
% Since $f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k$ is a continuous function of $\mathbf{x}_{-k}$ and the density $p(\mathbf{x}_{-k})$ is, by assumption, non-zero on the support, we conclude that for all $x_k, f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k =0 $ necessarily and thus for all $\mathbf{x}_{-k}$, $f(x_k, \mathbf{x}_{-k}) = f(0, \mathbf{x}_{-k}) + g(\mathbf{x}_{-k}) x_k$.

% The Hessian of $f$ then (guaranteed to exist by assumption) has a zero on the $k$-th main diagonal entry.

% By proposition 7.1.10 from Horn and Johnson~\cite{HJ90},  such a matrix is positive semidefinite if and only if the $k$-th row and column are also zero. (to prove this, one can first work with a 2 by 2 matrix and then use the fact that every principal submatrix of a PSD matrix must be PSD)

% Since $k$-th row and column correspond precisely to the gradient of $g(\mathbf{x}_{-k})$, we conclude that $g$ must be a constant function. It follows therefore that $g = 0$ because it integrates to 0.

% So we have that for all $x_k, \mathbf{x}_{-k}$, $f(x_k, \mathbf{x}_{-k}) = f(0, \mathbf{x}_{-k})$, which concludes our proof.

% \end{proof}

@@

%%% Local Variables: ***
%%% mode:latex ***
%%% TeX-master: "paper.tex" ***
%%% End: ***
